---
title: "UPCFLRP Data Exploration"
author: "Anson Call"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(ggpubr)
```

# Overview
While the broad goals of this project are outlined more completely in the [draft manuscript](https://cfri.app.box.com/file/1801500191498?s=j11vyr3c60lq4yxhi149anjlz9o1worc), the specific goals addressed here are to quantify treatment-associated changes in basal area, trees per acre, quadratic mean diameter (QMD), and species composition
- by treatment type (e.g., thinning, burning, etc.)
- by treatment intensity
- by treatment age or phase (e.g., 2-5 years post-treatment, 5-10 years post-treatment, etc.)
- by elevation and forest type (e.g., by unit)

# Concerns with specific plots

## Summary of pre-existing concerns

From Maggie's [Notes for (future) 2024 Data Analysis_MKP.docx](https://cfri.app.box.com/file/1445776069418):

> Lockhart plots might have overstory issues, it seems like some of them (L12, L23, L25, L13, L24 and L27) appear to have a DBH cutoff at 7.5-8 inch DBH in the PRE entry, but then show a large number of trees under 8 inches in DBH in post1 or post 5-6

> Be careful with plot 9D, some trees names RA9D, which I replaced with 9D for now but it’s possible they were duplicates.

In addition, the following 2013 plot visits are marked as "partial plots" in the ["UP-CFLRP All Plot Notes.docx"](https://cfri.app.box.com/file/1506842133331) 

```{r plots-marked-partial, echo=FALSE}
plots_marked_partial <- c("L12", "L13", "L23", "L24", "L25", "L27", "RA1510")
cat("Plots marked partial\n", plots_marked_partial)
```

The following two 2013 plot visits were marked with the note:

> Dropped in favor of 2015 pre

```{r additional-marked-plots, echo=FALSE}
addl_plots <- c("L29", "L8")
cat("Additional marked plots:\n", addl_plots)
```

## Confirming issues and identifying new ones

### Half-plots
From the [plot mapper](https://reshapewildfire.shinyapps.io/UPCFLRP_plot_mapper/), we can identify sites that have very high variation between visits. The following plots have a high coefficient of variation and a plot map suggesting that only half-plots were recorded on one or more visits:


```{r plot-mapper-problem-plots, echo=FALSE}
half_plots <- c(
  "RA261", # not actually a half plot?
  "RA1310" # confirmed half plot
  )

ntrees <- read.csv("UPCLRP_shiny_app/ntrees.csv")
ntrees %>%
  filter(id %in% half_plots)
```

The following plots are not exactly half-plots, but it's clear there is some incompleteness in the southeast quadrant:

```{r plot-mapper-problem-plots-continued, echo=FALSE}
partial_plots <- c(
  "RA219", # issue here is missing "RA216" data
  "RA2040", # issue here is extra trees (beyond )
  "RA710" # actually IS a half plot
  )

ntrees <- read.csv("UPCLRP_shiny_app/ntrees.csv")
ntrees %>%
  filter(id %in% partial_plots)
```

### Other issues
#### Inconsistent DBH thresholds
The following plots do not have evidence of the "half-plot" problem, but do have a high coefficient of variation, possibly because the lower DBH threshold for recording trees was not consistent across visits, as noted by Maggie:

```{r plot-mapper-problem-plots-continued_2, echo=FALSE}
addl_high_cv_plots <- c(
  "L29", "L7", "RA178", "L23", "3A", "RA181", "L8", "L9", "L27", "L25", "L11", 
  "RA281", "L10", "RA410", "L12", "L13", "L24"
)

ntrees <- read.csv("UPCLRP_shiny_app/ntrees.csv")
ntrees %>%
  filter(coef_variation > 0.2)
```

This list includes all of the plots identified by Maggie (L12, L23, L25, L13, L24 and L27) and adds 11 new ones. These new plots are not only from the "L" area. 

This problem of inconsistent lower DBH thresholds is most pronounced in these plots, but it actually seems like it could be *a problem in most, if not all* plots. 

We can examine this directly by checking the lower bounds of the DBH distribution for each plot and visit. We'll check the absolute minimum DBH and the 2.5th percentile of the DBH distribution for recorded trees at each plot and visit. Then, we can look for plots where the gap between these values on successive visits is large. 

```{r dbh-lower-bounds, message=FALSE, warning=FALSE}
overstory <- read_csv("overstory_tidy.csv")

min_dbh_change <- overstory %>%
  group_by(id, t_idxn) %>%
  summarise(
    min_dbh = min(dbh, na.rm = TRUE),
    p2_5_dbh = quantile(dbh, 0.025, na.rm = TRUE)
  ) %>%
  arrange(id, t_idxn) %>% 
  pivot_wider(
    names_from = t_idxn,
    values_from = c(min_dbh, p2_5_dbh)
  ) %>% 
  rowwise %>% 
  mutate(
    # What is the largest gap between the minimum DBH of the PRE_1 visit and all
    # POST_* visits?
    min_dbh_gap_PRE_1 = max(c_across((matches("^min_.*POST"))), na.rm = TRUE) - 
      min_dbh_PRE_1,
    # What is the largest gap between the minimum DBH of the PRE_2 visit and all
    # POST_* visits?
    min_dbh_gap_PRE_2 = max(c_across((matches("^min_.*POST"))), na.rm = TRUE) - 
      min_dbh_PRE_2,
    # same, but for 2.5%
    p2_5_dbh_gap_PRE_1 = max(c_across((matches("^p2_5_.*POST"))), na.rm = TRUE) - 
      p2_5_dbh_PRE_1,
    p2_5_dbh_gap_PRE_2 = max(c_across((matches("^p2_5_.*POST"))), na.rm = TRUE) - 
      p2_5_dbh_PRE_2
  ) %>% 
  filter(!all(is.na(val <- c_across(contains("gap"))) | val == -Inf)) %>%
  ungroup %>% 
  arrange(min_dbh_gap_PRE_1) %>% 
  mutate(row_highlight = id %in% addl_high_cv_plots,
         row_num = row_number())  # needed for row_spec()

min_dbh_change %>% 
  select(id, matches("gap"), starts_with("min"), starts_with("p2_5_")) %>% 
  kbl() %>%
  kable_paper() %>%
  column_spec(2:3, background = "#A1D5FF7F") %>% 
  column_spec(4:5, background = "#C7F9FF7F") %>% 
  row_spec(which(min_dbh_change$row_highlight), bold = TRUE) %>% 
  scroll_box(width = "100%", height = "300px")
```

It seems that the problems are all localized to PRE_1 visits. There isn't a big gap between the minimum values of the PRE_2 visits and any of the POST* visits. The problem plot values are bolded here, and you can see they are concentrated at the top of the table. That means a lot of the issue could be explained away by this gap in the size of trees that are being recorded.

Plot L7 is a big offender here:

```{r l7, echo=FALSE}
mins <- overstory %>% 
  filter(id == "L7") %>% 
  group_by(t_idxn) %>% 
  summarize(min = min(dbh))
overstory %>% 
  filter(id == "L7") %>% 
  ggplot(aes(x = dbh, fill = t_idxn)) +
  geom_histogram(alpha = 0.4, position = "identity") +
  geom_vline(xintercept = mins[[1, 2]], color = scales::hue_pal()(3)[1]) +
  geom_vline(xintercept = mins[[2, 2]], color = scales::hue_pal()(3)[2]) +
  geom_vline(xintercept = mins[[3, 2]], color = scales::hue_pal()(3)[3]) +
  labs(title = "Histogram of DBH values for plot L7, three total visits", 
       sub = "lines show minimum recorded dbh value")

```

Plotting histograms for all of the worst offenders **(watch out for changes in the color scales)**:

<div style="max-height: 600px; overflow-y: auto; width: 100%; border: 1px solid #ccc; padding: 10px;">
```{r dbh_histograms, echo=FALSE, results='hide', message=FALSE, warning=FALSE, class.source="scroll-800"}
# define plotting function
mk_dbh_hist <- \(ID, df) {
  mins <- df %>% 
    filter(id == ID) %>% 
    group_by(t_idxn) %>% 
    summarize(min = min(dbh))
  n_idxn <- nrow(mins)
  plot <- df %>% 
    filter(id == ID) %>% 
    ggplot(aes(x = dbh, fill = t_idxn)) +
    geom_histogram(alpha = 0.4, position = "identity") +
    geom_vline(xintercept = mins[[1, 2]], 
               color = scales::hue_pal()(n_idxn)[1]) +
    geom_vline(xintercept = mins[[2, 2]], 
               color = scales::hue_pal()(n_idxn)[2]) +
    labs(title = paste0("Histogram of DBH values for plot ",
                        ID,
                        " ",
                        n_idxn,
                        " total visits"), 
         sub = "lines show minimum recorded dbh value")
  if (n_idxn > 2) {
    plot <- plot + geom_vline(xintercept = mins[[3, 2]], 
                              color = scales::hue_pal()(n_idxn)[3])
  }
  if (n_idxn > 3) {
    plot <- plot + geom_vline(xintercept = mins[[4, 2]], 
                              color = scales::hue_pal()(n_idxn)[4]) 
  }
  return(plot)
}

# get ids to plot over
big_gaps <- min_dbh_change %>% 
  filter(min_dbh_gap_PRE_1 < -0.1) %>% 
  pull(id)
# make plots
plots <- map(big_gaps, \(x) mk_dbh_hist(x, overstory))
# arrange plots in scroll box
plots
```
</div>

A practical solution to this problem would be to _remove trees below the highest dbh threshold_ for plots where this seems to be an issue. 

```{r apply-cutoff}

overstory_min8 <- overstory %>% 
  filter(dbh >= 8)

data_loss <- (1 - nrow(overstory_min8) / nrow(overstory)) * 100

cat(round(data_loss, 2), "% of trees removed by DBH cutoff", sep = "")

```
Now, let's revisit the CV values for tree counts and see if some of these sites start to look less extreme.

```{r post-dbh-cv}
# ntrees data
ntrees_min8 <- overstory_min8 %>%
  group_by(id, t_idxn) %>%
  summarise(ntrees = n(), .groups = "keep") %>%
  ungroup() %>%
  pivot_wider(names_from = t_idxn, values_from = ntrees) %>%
  rowwise %>%
  mutate(mean = mean(c_across(starts_with("P")), na.rm = TRUE),
         sd = sd(c_across(starts_with("P")), na.rm = TRUE),
         coef_variation = ifelse(mean == 0, NA, sd / mean)) %>%
  select(id, starts_with("PRE"), starts_with("POST_"), starts_with("POSTRX"),
         mean, coef_variation) %>%
  arrange(desc(coef_variation)) %>% 
  ungroup %>% 
  mutate(cv_rank_new = row_number()) %>% 
  left_join(ntrees %>% 
              mutate(cv_rank_old = row_number(),
                     coef_variation_old = coef_variation) %>% 
              select(id, cv_rank_old, coef_variation_old), 
            by = "id") %>% 
  relocate(starts_with("cv_"), starts_with("coef_"), .before = PRE_1) %>% 
  arrange(cv_rank_old)

ntrees_min8 %>%
  kbl() %>%
  kable_paper() %>%
  column_spec(2:3, background = "#A1D5FF7F") %>% 
  column_spec(4:5, background = "#C7F9FF7F") %>% 
  row_spec(which(min_dbh_change$row_highlight), bold = TRUE) %>% 
  scroll_box(width = "100%", height = "800px") 
  

```

Okay, it looks like this correction clearly made a difference for some plots, particularly the `L` plots. The `RA` plot CVs are less affected. Let's glance at some histograms pre- and post-application of the DBH cutoff. 

<div style="max-height: 600px; overflow-y: auto; width: 100%; border: 1px solid #ccc; padding: 10px;">
```{r pre-post-histograms}
# define key plots: top 10 of ntrees_min8
key_plots <- ntrees_min8$id[1:10]
# map over key plots to make list of old plots
old_hists <- map(key_plots, \(x) mk_dbh_hist(x, overstory) + 
                   labs(subtitle = "OLD"))
new_hists <- map(key_plots, \(x) mk_dbh_hist(x, overstory_min8) + 
                   labs(subtitle = "NEW"))

str(new_hists)
dual_plot <- \(x, y) {
  plots <- list(x, y)
  p <- ggarrange(x, y, nrow = 2)
  p
}
dual_plot(old_hists[[1]], new_hists[[1]])
# plot old and new hists side-by-side
plots <- map2(old_hists, new_hists, \(x, y) dual_plot(x, y))
plots
```
</div>

#### Possible aspen regen

Though the DBH cutoff certainly helped, there are still some suspect cases of dramatically increasing tree counts in the post-treatment visits. One possible explanation is aspen regeneration. Let's see if aspen are frequent at any of the high-CV sites. 

```{r prop-aspen}

aspen_df <- overstory_min8 %>%
  group_by(id, species) %>%
  summarise(ntrees = n(), .groups = "keep") %>%
  ungroup() %>%
  pivot_wider(names_from = species, values_from = ntrees) %>%
  rowwise %>%
  mutate(aspen_proportion = POTR5 / sum(across(PIPO:last_col()), 
                                        na.rm = TRUE)) %>% 
  select(id, aspen_proportion)

out <- ntrees_min8 %>%
  left_join(aspen_df, by = "id") %>% 
  relocate(aspen_proportion, .before = everything()) %>% 
  arrange(cv_rank_new) %>% 
  mutate(aspen_perc_rank = percent_rank(aspen_proportion), .before = id)

out %>% 
  kbl() %>%
  kable_paper() %>%
  column_spec(1:2, background = "#A1D5FF7F") %>% 
  column_spec(4, background = "#C7F9FF7F") %>% 
  row_spec(which(out$aspen_perc_rank > 0.75), bold = TRUE) %>% 
  scroll_box(width = "100%", height = "800px") 
```

# TODO
  - address maggie's email: some plots (e.g. RA261 and RA2610) may be improperly lumped together in `overstory.R`
  - stump data
  - lay out full diagnosis and recommend and solution for each plot

#### Stump data

Finally, I checked to see which plots had stump data, because I think these are the plots that were of concern previously. 

```{r stump-dbh, echo=FALSE}
overstory_raw <- suppressWarnings(
  readxl::read_excel("raw/RA_data_Master_20250224.xlsx", 
                                    sheet = "Overstory")
  )
overstory_raw %>% 
  filter(Status == "S" & !is.null(DBH)) %>% 
  pull(PLOT) %>% 
  unique()
```

I'm seeing another issue: the plot IDs do not seem to be recorded consistently. _2025/07/22 update:_ this has now been fixed in overstory.R.\n See the `# check id ####` section. 

Additionally, there are lots of plots which have only one site visit:


## Overview plots 

## Some simple statistical analyses to support the plots

```{r cars}
summary(cars)
```
